{\boldmath Pseudoscalar top-Higgs coupling: Exploration of $\mathrm{CP}$-odd observables to resolve the sign ambiguity }

### 2016-03-03 Paper_ttbarH_Draft.tex
KK edits through Sec. V

The mean values of $\epsilon_8$ for the scenarios under
consideration are comparable with the values listed in Table
\ref{table6} for $\epsilon_5$. Concerning the associated angular
distributions, their range of variation {\bf (KK: I'm not sure what
  this means)} is larger than in the case of
the distributions related to $\epsilon_5$-$\epsilon_7$ but exhibit
smaller changes in their shapes for the different scenarios and for
this reason we do not include here the corresponding plots.
{\bf (KK: might need to reword this a bit after clarifying what 
  range of variation means.)}
\caption{Asymmetry for the TP $\epsilon_{8}$ for the SM case and the
  two $\mathrm{CP}$-mixed cases defined by $\kp=1,\kpt=\pm 1$. The
  values are obtained with $10^5$ simulated events.}
\label{table8}
\begin{tabular}{|C{1cm}|C{1cm}||C{2cm}|C{2cm}|}
\hhline{|====|}
$\kappa_t$&$\tilde{\kappa}_t$~~&$\mathcal{A}(\epsilon_8)$~~&$\mathcal{A}(\epsilon_8)/\sigma_{\mathcal{A}}$ \\ 
\hhline{|====|} 
$1$ & $-1$~~~ & $0.0331$~ & $10.5$~ \\[0.6mm]
$1$ & $0$ & $0.0023$~ & $0.7$~ \\[0.6mm]
$1$ & $1$ & $-0.0403$~~~ & $\,-12.7$~~~~ \\[0.6mm]
\hhline{|====|}
\section{{\boldmath{$\mathrm{CP}$}}-odd observables not depending on \MakeLowercase{{\boldmath $t$}} and \MakeLowercase{{\boldmath $\tbar$}} momenta}
\label{sec5}
All the observables discussed in the above sections involve the
momenta of the top and anti-top quarks thus requiring the full
reconstruction of the kinematics of the individual $t$ and $\tbar$
systems in order to be measured. Although challenging due to the
presence of two neutrinos in the final state, this can be in principle
done by applying a kinematic reconstruction method such that the
neutrino weighting technique \cite{atlasconf,atlascharge}. Another
possibility is to define observables that are not dependent on the $t$
and $\tbar$ momenta but make use of the quarks $b$ and $\bbar$. In
order to construct such observables we will modify here the most
sensitive observables studied in Secs.~\ref{table3} and \ref{sec4},
namely the combinations $\epsilon_4$ and $\epsilon_8$
respectively. Let us first consider the combination $\epsilon_8$
defined in Eq.~(\ref{eq32}) and replace the top and antitop momenta by
the bottom and antibottom momenta respectively. Thus, we define
\label{eq33}
\epsilon_9 = \epsilon(b+\bbar+H,b-\bbar,\ell^+,\ell^-).
Note that now the sign of the TP $\epsilon_9$ is set by the sign of
the quantity $(\vec{b}-\vec{\bbar})\cdot(\vec{\ell}^+\!\times
\vec{\ell}^-)$ in the $b\bbar H$ rest frame. This quantity is in fact
used in \cite{Guadagnoli} but in the lab frame in order to define a
$\mathrm{CP}$-odd observable that only depends on lab frame
variables. The values of the asymmetry for $\epsilon_9$ are listed in
Table \ref{table9}. By comparing Tables \ref{table8} and \ref{table9}
we see that the use of the $b,\bbar$ instead of $t,\tbar$ leads to a
decrese in the sensitivity of the asymmetry by $\sim 5\sigma$ for
$\kp=1,\kpt=\pm 1$. However, the observable is still capable of
discriminating not only between the two $\mathrm{CP}$-mixed hypotheses
but also between these and the SM case.
%\newpage
We proceed now in a similar way with the combination $\epsilon_4$. By
using Eq.~(\ref{eq24}) along with the definitions of the spin vectors
in Eqs.~(\ref{eq3})-(\ref{eq4}), we can write
\label{eq34}
\epsilon_4 = \frac{m^2_t}{(t\cdot\ell^+)\cdot(\tbar\cdot\ell^-)}\,\epsilon(Q,t-\tbar,\ell^-,\ell^+)+\frac{1}{(t\cdot\ell^+)}\,\epsilon(Q,t,\ell^+,\tbar)-\frac{1}{(\tbar\cdot \ell^-)}\,\epsilon(Q,\tbar,t,\ell^-).
Since the asymmetry is not changed by an overall positive definite
factor, we will concentrate on the following combination arising from
the expression in Eq.~(\ref{eq34}),
\label{eq35}
\epsilon(Q,t-\tbar,\ell^-,\ell^+)+\frac{(\tbar\cdot \ell^-)}{m^2_t}\epsilon(Q,t,\ell^+,\tbar)-\frac{(t\cdot\ell^+)}{m^2_t}\epsilon(Q,\tbar,t,\ell^-),
\caption{Asymmetry for the TP $\epsilon_{9}$ for the SM case and the
  two $\mathrm{CP}$-mixed cases defined by $\kp=1,\kpt=\pm 1$. The
  values are obtained with $10^5$ simulated events.}
\label{table9}
\begin{tabular}{|C{1cm}|C{1cm}||C{2cm}|C{2cm}|}
\hhline{|====|}
$\kappa_t$&$\tilde{\kappa}_t$~~&$\mathcal{A}(\epsilon_9)$~~&$\mathcal{A}(\epsilon_9)/\sigma_{\mathcal{A}}$ \\ 
\hhline{|====|} 
$1$ & $-1$~~~ & $0.0171$~ & $5.4$~ \\[0.6mm]
$1$ & $0$ & $0.0010$~ & $0.3$~ \\[0.6mm]
$1$ & $1$ & $-0.0247$~~~ & $\,-7.8$~~~~ \\[0.6mm]
\hhline{|====|}
and instead of replacing $t$ and $\tbar$ directly by $b$ and $\bbar$,
we use their visible parts, namely $b+\ell^+$ and $\bbar +\ell^-$
respectively. This results in the following definition
\label{eq36}
\epsilon_{10}=\epsilon(\tilde{Q},c_{b\bbar}\,,\ell^-,\ell^+)-w_1\,\epsilon(\tilde{Q},b,\bbar,\ell^+)+w_2\,\epsilon(\tilde{Q},b,\bbar,\ell^-),
where $\tilde{Q}\equiv (b+\ell^+\!+\bbar +\ell^-)/2$ stands for the
visible part of $Q$, $c_{b\bbar}=(1-w_1)\,b-(1-w_2)\,\bbar$ and the
weights $w_{1,2}$ are given by $(\bbar\cdot \ell^-)/m^2_t $ and
$(b\cdot \ell^+)/m^2_t$ respectively. Also, the contribution
$m^2_{\ell}/m^2_t$ has been neglected both in $w_1$ and in $w_2$. Note
that if we set $w_1=w_2=0$, the combination $\epsilon_{10}$ reduces to
$\epsilon_9 /2$ and $\mathcal{A}(\epsilon_{10})$ becomes equal to
$\mathcal{A}(\epsilon_9)$. The results obatined for the asymmetry of
$\epsilon_{10}$ are given in Table \ref{table10}. By comparing Tables
\ref{table2} and \ref{table10} we see again that the sensitivity of
the asymmetry decreases when $t$ and $\tbar$ are not included in the
TP. Nevertheless, the combination $\epsilon_{10}$ remains a useful
observable for discriminating the $\mathrm{CP}$ nature of the Higgs
boson, with the corresponding asymmetry having a sensitivity even
higher than that of $\epsilon_9$, as can be checked by taking into
account Tables \ref{table9} and \ref{table10}. More precisely, the
separation between the $\mathrm{CP}$-mixed hypotheses is enhanced by
about $3\sigma$. This improvement in the asymmetry of $\epsilon_{10}$
with respect to $\epsilon_9$ may be due to two facts. In the first
place, as was pointed out in Sec.~\ref{sec4} when comparing the TPs
$\epsilon_4$ and $\epsilon_8$, the asymmetry appears to be higher when
the spin vectors are used instead of the lepton momenta and we see
from Eqs.~(\ref{eq33}) and (\ref{eq36}) that $\epsilon_{10}$, being
obtained from $\epsilon_4$, contains the information on the spin
vectors, as opposed to $\epsilon_9$ that depends directly on the
lepton momenta because it is derived from $\epsilon_8$. In the second
place, in order to obtain $\epsilon_{10}$ we have replaced in the
expression for $\epsilon_4$ the top and antitop momenta by their
visible part, while in the case of $\epsilon_9$ the bottom and
antibottom momenta have been used.
\caption{Asymmetry for the TP $\epsilon_{10}$ for the SM case and the
  two $\mathrm{CP}$-mixed cases defined by $\kp=1,\kpt=\pm 1$. The
  values are obtained by using $10^5$ simulated events.}
\label{table10}
\begin{tabular}{|C{1cm}|C{1cm}||C{2cm}|C{2cm}|}
\hhline{|====|}
$\kappa_t$&$\tilde{\kappa}_t$~~&$\mathcal{A}(\epsilon_{10})$~~&$\mathcal{A}(\epsilon_{10})/\sigma_{\mathcal{A}}$ \\ 
\hhline{|====|} 
$1$ & $-1$~~~ & $-0.0213$~~~ & $-6.7$~~~ \\[0.6mm]
$1$ & $0$ & $0.0031$~ & $1.0$~ \\[0.6mm]
$1$ & $1$ & $0.0300$~ & $9.5$~ \\[0.6mm]
\hhline{|====|}
By using our set of simulated events we have also tested, for
comparison purposes, the asymmetry of the lab frame observable given
in \cite{Guadagnoli}. We have obtained that this observable appears to

@@

giving rise to a separation between the $\mathrm{CP}$-mixed hypotheses
smaller by about $1.4\sigma$.

>>>>

The mean values of $\epsilon_8$ for the scenarios under
consideration are comparable with the values listed in Table
\ref{table6} for $\epsilon_5$. Concerning the associated angular
distributions, their range of variation {\bf (KK: I'm not sure what
  this means)} is larger than in the case of
the distributions related to $\epsilon_5$-$\epsilon_7$ but exhibit
smaller changes in their shapes for the different scenarios and for
this reason we do not include here the corresponding plots.
{\bf (KK: might need to reword this a bit after clarifying what 
  range of variation means.)}
\begin{table}[H]
\caption{Asymmetry for the TP $\epsilon_{8}$ for the SM case and the
  two $\mathrm{CP}$-mixed cases defined by $\kp=1,\kpt=\pm 1$. The
  values are obtained with $10^5$ simulated events.}
\label{table8}
\begin{center}
\begin{tabular}{|C{1cm}|C{1cm}||C{2cm}|C{2cm}|}
%\begin{tabular}{|c|r||r|c||r|c||r|c|}
\hhline{|====|}
%\hhline{|--------|}
$\kappa_t$&$\tilde{\kappa}_t$~~&$\mathcal{A}(\epsilon_8)$~~&$\mathcal{A}(\epsilon_8)/\sigma_{\mathcal{A}}$ \\ 
\hhline{|====|} 
%\hhline{|--------|}
$1$ & $-1$~~~ & $0.0331$~ & $10.5$~ \\[0.6mm]
\hline
$1$ & $0$ & $0.0023$~ & $0.7$~ \\[0.6mm]
\hline
$1$ & $1$ & $-0.0403$~~~ & $\,-12.7$~~~~ \\[0.6mm]
\hhline{|====|}
%\hhline{|--------|}
\end{tabular}
\end{center} 
\end{table}
\section{{\boldmath{$\mathrm{CP}$}}-odd observables not depending on \MakeLowercase{{\boldmath $t$}} and \MakeLowercase{{\boldmath $\tbar$}} momenta}
\label{sec5}
The observables discussed in the preceding sections all involve the
momenta of the top and/or anti-top quarks and thus require the full
reconstruction of the kinematics of the individual $t$ and $\tbar$
systems in order to be measured. Although challenging due to the
presence of the two neutrinos in the final state, this can in principle be
done by applying a kinematic reconstruction method such as the
neutrino weighting technique~\cite{atlasconf,atlascharge}. Another
possibility is to define observables that do not depend on the $t$
and $\tbar$ momenta but instead make use of the momenta of the
$b$ and $\bbar$ quarks to which the $t$ and $\tbar$ decay. In
order to construct such observables we will take as our starting
point the most
sensitive observables studied in Secs.~\ref{sec3} and \ref{sec4},
{\bf (KK: note that I changed the reference from table3 to sec3.)}
namely those associated with the TPs $\epsilon_4$ and $\epsilon_8$,
respectively.

Let us first consider the TP combination $\epsilon_8$, which is
defined in Eq.~(\ref{eq32}).  Replacing the momenta of the
$t$ and $\tbar$ quarks by the momenta of the $b$ and $\bbar$ quarks,
respectively, we have a new TP,
%
\beq
\label{eq33}
\epsilon_9 = \epsilon(b+\bbar+H,b-\bbar,\ell^+,\ell^-).
\eeq
%
Note that the sign of $\epsilon_9$ is determined by the sign of
the quantity $(\vec{b}-\vec{\bbar})\cdot(\vec{\ell}^+\!\times
\vec{\ell}^-)$ in the $b\bbar H$ rest frame. This combination
of three vectors (determined in the lab frame instead
of the $b\bbar H$ rest frame) is
used in Ref.~\cite{Guadagnoli} to define a
$\mathrm{CP}$-odd observable that only depends on lab frame
variables. The values of the asymmetry for $\epsilon_9$ are listed in
Table~\ref{table9}. Comparing Tables~\ref{table8} and \ref{table9}
we see that the use of the $b$ and $\bbar$ momenta
instead of the $t$ and $\tbar$ momenta leads to a
decrease in the sensitivity of the asymmetry by $\sim 5\sigma$ for
$\kp=1,\kpt=\pm 1$. Nevertheless, the observable can still
discriminate not only between the two $\mathrm{CP}$-mixed scenarios
but also between these and the SM case.
%\newpage

%
\begin{table}[th]
\caption{Asymmetry for the TP $\epsilon_{9}$ for the SM case and the
  two $\mathrm{CP}$-mixed cases defined by $\kp=1,\kpt=\pm 1$. The
  values are obtained with $10^5$ simulated events.
  {\bf (KK: I repositioned the table in the file, please double check to make
  sure I didn't mess anything up :)}}
\label{table9}
\begin{center}
\begin{tabular}{|C{1cm}|C{1cm}||C{2cm}|C{2cm}|}
%\begin{tabular}{|c|r||r|c||r|c||r|c|}
\hhline{|====|}
%\hhline{|--------|}
$\kappa_t$&$\tilde{\kappa}_t$~~&$\mathcal{A}(\epsilon_9)$~~&$\mathcal{A}(\epsilon_9)/\sigma_{\mathcal{A}}$ \\ 
\hhline{|====|} 
%\hhline{|--------|}
$1$ & $-1$~~~ & $0.0171$~ & $5.4$~ \\[0.6mm]
\hline
$1$ & $0$ & $0.0010$~ & $0.3$~ \\[0.6mm]
\hline
$1$ & $1$ & $-0.0247$~~~ & $\,-7.8$~~~~ \\[0.6mm]
\hhline{|====|}
%\hhline{|--------|}
\end{tabular}
\end{center} 
\end{table}
%

We proceed in a similar manner with the TP $\epsilon_4$. Starting from
Eq.~(\ref{eq24}) and using the definitions of the spin vectors
in Eqs.~(\ref{eq3})-(\ref{eq4}), we have
%
\beq
\label{eq34}
\epsilon_4 = \frac{m^2_t}{(t\cdot\ell^+)\cdot(\tbar\cdot\ell^-)}\,\epsilon(Q,t-\tbar,\ell^-,\ell^+)+\frac{1}{(t\cdot\ell^+)}\,\epsilon(Q,t,\ell^+,\tbar)-\frac{1}{(\tbar\cdot \ell^-)}\,\epsilon(Q,\tbar,t,\ell^-).
\eeq
%
Since the asymmetry is not changed by the presence
of an overall positive definite multiplicative
factor, let us concentrate instead on the following combination
of TPs,
%
\beq
\label{eq35}
\epsilon(Q,t-\tbar,\ell^-,\ell^+)+\frac{(\tbar\cdot \ell^-)}{m^2_t}\epsilon(Q,t,\ell^+,\tbar)-\frac{(t\cdot\ell^+)}{m^2_t}\epsilon(Q,\tbar,t,\ell^-).
\eeq
Instead of replacing $t$ and $\tbar$ directly by $b$ and $\bbar$,
we use the visible contributions, namely $b+\ell^+$ and $\bbar +\ell^-$,
respectively. This results in the following definition
%
\beq
\label{eq36}
\epsilon_{10}=\epsilon(\tilde{Q},c_{b\bbar}\,,\ell^-,\ell^+)-w_1\,\epsilon(\tilde{Q},b,\bbar,\ell^+)+w_2\,\epsilon(\tilde{Q},b,\bbar,\ell^-),
\eeq
%
where $\tilde{Q}\equiv (b+\ell^+\!+\bbar +\ell^-)/2$ stands for the
visible part of $Q$, $c_{b\bbar}=(1-w_1)\,b-(1-w_2)\,\bbar$, and the
weights $w_{1,2}$ are given by $(\bbar\cdot \ell^-)/m^2_t $ and
$(b\cdot \ell^+)/m^2_t$, respectively. Also, the contribution
$m^2_{\ell}/m^2_t$ has been neglected both in $w_1$ and in $w_2$. Note
that if we set $w_1=w_2=0$, the combination $\epsilon_{10}$ reduces to
$\epsilon_9 /2$ and $\mathcal{A}(\epsilon_{10})$ becomes equal to
$\mathcal{A}(\epsilon_9)$. The results obtained for the asymmetry of
$\epsilon_{10}$ are given in Table \ref{table10}. By comparing Tables
\ref{table2} and \ref{table10} we see again that the sensitivity of
the asymmetry decreases when $t$ and $\tbar$ are not included in the
TP. Nevertheless, the combination $\epsilon_{10}$ remains a useful
observable for discriminating the $\mathrm{CP}$ nature of the Higgs
boson, with the corresponding asymmetry having a sensitivity that is
higher than that of $\epsilon_9$.

Comparing
Tables~\ref{table9} and \ref{table10}, we see that the
separation between the $\mathrm{CP}$-mixed scenarios is enhanced by
about $3\sigma$ for $\mathcal{A}(\epsilon_{10})$ compared
  to $\mathcal{A}(\epsilon_{9})$.
    This improvement in the asymmetry may be due to two facts. In the first
place, as was pointed out in Sec.~\ref{sec4} when comparing the TPs
$\epsilon_4$ and $\epsilon_8$, the asymmetry appears to be higher when
the spin vectors are used instead of the lepton momenta.  We see
from Eqs.~(\ref{eq33}) and (\ref{eq36}) that $\epsilon_{10}$, being
obtained from $\epsilon_4$, contains the information on the spin
vectors; by way of contrast, $\epsilon_9$ depends directly on the
lepton momenta because it is derived from $\epsilon_8$. In the second
place, in order to obtain $\epsilon_{10}$, we have replaced
the top and antitop momenta by their 
visible parts, while in the case of $\epsilon_9$ the bottom and
antibottom momenta have been used.

%
\begin{table}[th]
\caption{Asymmetry for the TP $\epsilon_{10}$ for the SM case and the
  two $\mathrm{CP}$-mixed cases defined by $\kp=1,\kpt=\pm 1$. The
  values are obtained by using $10^5$ simulated events.}
\label{table10}
\begin{center}
\begin{tabular}{|C{1cm}|C{1cm}||C{2cm}|C{2cm}|}
%\begin{tabular}{|c|r||r|c||r|c||r|c|}
\hhline{|====|}
%\hhline{|--------|}
$\kappa_t$&$\tilde{\kappa}_t$~~&$\mathcal{A}(\epsilon_{10})$~~&$\mathcal{A}(\epsilon_{10})/\sigma_{\mathcal{A}}$ \\ 
\hhline{|====|} 
%\hhline{|--------|}
$1$ & $-1$~~~ & $-0.0213$~~~ & $-6.7$~~~ \\[0.6mm]
\hline
$1$ & $0$ & $0.0031$~ & $1.0$~ \\[0.6mm]
\hline
$1$ & $1$ & $0.0300$~ & $9.5$~ \\[0.6mm]
\hhline{|====|}
%\hhline{|--------|}
\end{tabular}
\end{center} 
\end{table}
%

For comparison purposes, we have also used our simulated events
to test the asymmetry corresponding to the lab frame observable given
in Ref.~\cite{Guadagnoli}. {\bf (KK: Is this lab frame observable
  just a TP, like we have used?  It seems strange to talk about an
  asymmetry associated with an observable, when normally an asymmetry
  is itself considered to be an observable.  I'm wondering if we are
  considering an asymmetry associated with a TP, or if we are actually
  just looking at whatever observable they have defined\ldots.  Sorry for
  my confusion on this!)}  We have found that this observable
{\bf (KK: ``asymmetry''?)}
appears to

@@

{\bf (KK: ``$\mathcal{A}(\epsilon_{10})$''?)}
giving rise to a separation between the $\mathrm{CP}$-mixed scenarios
that is smaller by about $1.4\sigma$.
